---
title: "Affymetrix data analysis"
author: "Ignasi Jarne Sanz"
date: "19/4/2022"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Working directory and packages
````{r}
setwd("C:/Users/User/Desktop/omic/")
````

`````{r warning = FALSE, message= FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
library(oligo)
library(Biobase)
library(arrayQualityMetrics)
library(ggplot2)
library(ggrepel)
library(Rcpp)
library(genefilter)
library(clariomdhumantranscriptcluster.db)
library(limma)
library(org.Hs.eg.db)
library(grDevices)
library(ggnewscale)
library(gplots)
library(GO.db)
library(GOstats)
library(reactome.db)
library(ReactomePA)
`````

## Characterization and sample selection.
#### Aleatory sample selection
Out of 47 total samples, we will collect 15 (5 per group). 
This 47 samples are stored in a subdirectory called "copia".
````{r}
noms_raw <- list.files("./copia/")
numbers <- seq(1,47, by=1)
fenotip <- rep(c("Symptomatic","Asymptomatic","Healthy controls"),
               times=c(11,18,18))

Muestras <- data.frame(numbers,noms_raw,fenotip)
head(Muestras)

set.seed(2907)
Sym <- sample(1:11, 5, replace = F);sort(Sym)
Asym <- sample(12:29, 5, replace = F);sort(Asym)
Heal <- sample(30:47, 5, replace = F);sort(Heal)
print(Muestras[c(Sym,Asym,Heal),])
````

We create a target tile based on the selected samples, this target file is stored in a subdirectory called "dades".
````{r, include=FALSE}
targets <- read.csv2("./dades/targets.csv")

targets %>%
  kbl(caption = "Selected samples") %>%
  kable_classic(full_width = T, html_font = "Cambria")
````

We list the selected files, that we will have stored a subdirectory called "dades".
````{r warinig=FALSE, message = FALSE}
celFiles <- list.celfiles("./dades", full.names = T)
my.targets <- read.AnnotatedDataFrame(file.path("./dades","targets.csv"),
                                      header = T, row.names = 1,
                                      sep = ";")
rawData <- read.celfiles(celFiles, phenoData = my.targets)
````

Change the data names for readable names.
````{r}
head(rawData@assayData[["exprs"]])
colnames(rawData) <-rownames(pData(rawData)) <- my.targets@data$ShortName # Changing the name
head(rawData@assayData[["exprs"]]) # Checking the names
````


## Quality control and data exploration of the raw expression data
Applying "arrayQualitMetrics" on the raw data, this creates a .html file with quality metrics regardin the loaded data.
````{r, warning=FALSE}
arrayQualityMetrics(rawData,
                    outdir =file.path("./resultats", "QC_Raw"), force = T)
````

Some exploratory analysis on our expression data

### Density plot
````{r}
colors <- rep(c(1,2,3),times=c(5,5,5))
hist(rawData, main="Signal distribution for the raw expression data", col=colors)
````

### Boxplots
````{r}
boxplot(rawData, main="Signal distribution for the raw expression data", col=colors, las=2)
````

Las g치ficas generadas en este apartado nos muestran que los datos no est치n normalizados.

### PCA
````{r}
# funci칩n
plotPCA3 <- function (datos, labels, factor, title, scale,colores, size = 1.5, glineas = 0.25) {
  data <- prcomp(t(datos),scale=scale)
  # plot adjustments
  dataDf <- data.frame(data$x)
  Group <- factor
  loads <- round(data$sdev^2/sum(data$sdev^2)*100,1)
  # main plot
  p1 <- ggplot(dataDf,aes(x=PC1, y=PC2)) +
    theme_classic() +
    geom_point(aes(color = Group), alpha = 0.55, size = 3) +
    coord_cartesian(xlim = c(min(data$x[,1])-5,max(data$x[,1])+5)) +
    scale_fill_discrete(name = "Group")
  p1 + geom_text(aes(y = PC2 +6500, x=PC1+15000,  
                     label = labels), size=size)+
    labs(x = c(paste("PC1",loads[1],"%")),y=c(paste("PC2",loads[2],"%"))) +  
    ggtitle("PCA for the raw data")+ 
    theme(plot.title = element_text(hjust = 0.5)) +
    scale_color_manual(values=colores)
}

plotPCA3(exprs(rawData), labels=targets$ShortName, factor = targets$Grup,
        title = "Raw Data", scale=FALSE, size=3,
        colores = c("red","blue","green"))
````


## Expression data RMA transformation
````{r}
normData <- rma(rawData)
````

## Quality control on the RMA transformed data
````{r warning=FALSE}
arrayQualityMetrics(normData,
                    outdir =file.path("./resultats", "QC_Norm"), force = T)
````

### Density plot
````{r}
colors <- rep(c(1,2,3),times=c(5,5,5))
hist(normData, main="Signal distribution for the normalized expression data", col=colors)
````

### Boxplots
````{r}
boxplot(normData, main="Signal distribution for the normalized expression data", col=colors, las=2)
````

### PCA
````{r}
# funci칩n
plotPCA3 <- function (datos, labels, factor, title, scale,colores, size = 1.5, glineas = 0.25) {
  data <- prcomp(t(datos),scale=scale)
  # plot adjustments
  dataDf <- data.frame(data$x)
  Group <- factor
  loads <- round(data$sdev^2/sum(data$sdev^2)*100,1)
  # main plot
  p1 <- ggplot(dataDf,aes(x=PC1, y=PC2)) +
    theme_classic() +
    geom_point(aes(color = Group), alpha = 0.55, size = 3) +
    coord_cartesian(xlim = c(min(data$x[,1])-5,max(data$x[,1])+5)) +
    scale_fill_discrete(name 
                        = "Group")
  # avoiding labels superposition
  p1 + geom_text(aes(y = PC2+5, x=PC1+5,  
                     label = labels), size=size)+
    labs(x = c(paste("PC1",loads[1],"%")),y=c(paste("PC2",loads[2],"%"))) +  
    ggtitle("PCA for the normalized data")+
    theme(plot.title = element_text(hjust = 0.5)) +
    scale_color_manual(values=colores)
}

plotPCA3(exprs(normData), labels=targets$ShortName, factor = targets$Grup,
        title = "Norm", scale=FALSE, size=3,
        colores = c("red","blue","green"))
````


## Non specific gene filtering
We will filter the 25% of genes that show the greatest variation between groups. This step is not mandatory.
````{r}
annotation(normData) <- "clariomdhumantranscriptcluster.db"
filtered <- nsFilter(normData,
                     require.entrez = T, remove.dupEntrez = T,
                     var.filter = T, var.func = IQR, var.cutoff = 0.75,
                     filterByQuantile = T, feature.exclude = "^AFFX")

print(filtered$filter.log)
````


```{r}
filtered_data <- filtered$eset
filteredData <- exprs(filtered_data)
colnames(filteredData) <- pData(filtered$eset)$ShortName
```


## DEGs identification
Design matrix
```{r}
grup <- pData(filtered$eset)$Grup
lev <- factor(grup, levels = unique(grup))
design <- model.matrix(~0+lev)
colnames(design) <- levels(lev)
rownames(design) <- targets$ShortName
print(design)
```

Contrast matrix
```{r}
contrast_mat <- makeContrasts(SympvsAsymp = Symp-Asymp,
                              SympvsHealth = Symp-Healthy,
                              AsympvsHealth = Asymp-Healthy,
                              levels = design)
print(contrast_mat)
```

Linear model generation
```{r}
fit <- lmFit(filteredData, design)
fiit.main <- contrasts.fit(fit, contrast_mat)
fiit.main <- eBayes(fiit.main)
```

Results extraction
```{r}
TT_Sym_vs_Asym <- topTable(fiit.main, number = nrow(fiit.main), coef = "SympvsAsymp",adjust.method = "fdr")
TT_Sym_vs_Health <- topTable(fiit.main, number = nrow(fiit.main), coef = "SympvsHealth",adjust.method = "fdr")
TT_Asymp_vs_Health <- topTable(fiit.main, number = nrow(fiit.main), coef = "AsympvsHealth",adjust.method = "fdr")
```

Some numbers
```{r}
DEGs1 <- nrow(topTable(fiit.main, number = nrow(fiit.main), coef = "SympvsAsymp",adjust.method = "fdr",
                       lfc = 2, p.value = 0.05));DEGs1
DEGs2 <- nrow(topTable(fiit.main, number = nrow(fiit.main), coef = "SympvsHealth",adjust.method = "fdr",
                       lfc = 2, p.value = 0.05));DEGs2
DEGs3 <- nrow(topTable(fiit.main, number = nrow(fiit.main), coef = "AsympvsHealth",adjust.method = "fdr",
                       lfc = 2, p.value = 0.05));DEGs3
```


## Results annotations
````{r}
keytypes(clariomdhumantranscriptcluster.db)
````
We will extract the following information:

- Symbol

- ENTREZID

- Gene name

````{r}
annotations <- AnnotationDbi::select(clariomdhumantranscriptcluster.db, keys = rownames(filteredData),columns = c("SYMBOL","GENENAME","ENTREZID"))
head(annotations)
````

Getting the extracted annotation with our results.
````{r}
TT_Sym_vs_Asym <- TT_Sym_vs_Asym %>%
  mutate(PROBEID=rownames(TT_Sym_vs_Asym)) %>%
  left_join(annotations) %>%
  arrange(P.Value)
TT_Sym_vs_Asym <- TT_Sym_vs_Asym[,c(7,8,9,10,
                                    1,2,3,4,5,6)]


TT_Sym_vs_Health <- TT_Sym_vs_Health %>%
  mutate(PROBEID=rownames(TT_Sym_vs_Health)) %>%
  left_join(annotations) %>%
  arrange(P.Value)
TT_Sym_vs_Health <- TT_Sym_vs_Health[,c(7,8,9,10,
                                    1,2,3,4,5,6)]



TT_Asymp_vs_Health <- TT_Asymp_vs_Health %>%
  mutate(PROBEID=rownames(TT_Asymp_vs_Health)) %>%
  left_join(annotations) %>%
  arrange(P.Value)
TT_Asymp_vs_Health <- TT_Asymp_vs_Health[,c(7,8,9,10,
                                    1,2,3,4,5,6)]

head(TT_Sym_vs_Asym)
````

## Comparing results.
````{r}
gene_sym <- select(clariomdhumantranscriptcluster.db, rownames(fiit.main), c("SYMBOL"))
gene_sym <- gene_sym$SYMBOL
````

### Volcano plots
#### Contrast 1
````{r}
volcanoplot(fiit.main, coef = 1, highlight = 15, names = gene_sym,main=paste("DEGs:", colnames(contrast_mat)[1]))
````

#### Contrast 2
````{r}
volcanoplot(fiit.main, coef = 2, highlight = 15, names = gene_sym,main=paste("DEGs:", colnames(contrast_mat)[2]))
````

#### Contrast 3
````{r}
volcanoplot(fiit.main, coef = 3, highlight = 15, names = gene_sym,main=paste("DEGs:", colnames(contrast_mat)[3]))
````

### Comparing results
#### Venn Diagram
Preparing the data
````{r}
comparing <- decideTests(fiit.main, method = "separate", adjust.method = "fdr",
                         p.value = 0.05, lfc = 2)
comparing_rows <- apply(abs(comparing),1,sum)
comparing_sel <- comparing[comparing_rows!=0,]
print(summary(comparing))
````

Plotting
```{r}
vennDiagram(comparing_sel[,1:3], cex = 0.9,
            circle.col = c("blue","red","green"))
title("Common DEGs between contrasts\n (p.adj < 0.05 y LFC>2)")
```


#### Heatmap
Preparing the data
````{r}
probes_heat <- rownames(comparing_sel)
heat_data <- exprs(filtered_data)[rownames(exprs(filtered_data)) %in% probes_heat,]
heat_symbols <- select(clariomdhumantranscriptcluster.db, rownames(heat_data), c("SYMBOL"))
heat_symbols <- heat_symbols$SYMBOL
rownames(heat_data) <- heat_symbols
head(heat_data)
````

Plotting
````{r}
heatmap.2(heat_data,
          Rowv = T,
          Colv = T,
          dendrogram = "both",
          main = "Genes diferencialmente expresados\n (FDR <0.05 lFC >2)",
          scale = "row",
          col = colorRampPalette(c("blue","red"))(n=300),
          sepcolor = "white",
          sepwidth = c(0.05,0.05),
          cexRow = 0.5,
          cexCol = 0.9,
          key = T,
          keysize = 1.5,
          density.info = "density",
          ColSideColors = c(rep("red",5),rep("blue",5),rep("green",5)),
          tracecol = NULL,
          srtCol = 30)
````

## Biological significance analysis
Preparing the data
````{r}
tables_to_use <- list(Symp_vs_Asymp = TT_Sym_vs_Asym,
                      Symp_vs_Healthy = TT_Sym_vs_Health,
                      Asym_vs_Healty = TT_Asymp_vs_Health)

list_of_selected <- list()
for (i in 1:length(tables_to_use)){
  topTab <- tables_to_use[[i]]
  genes_to_use <- topTab["adj.P.Val"]<0.15
  IDs_to_use <- (topTab$PROBEID)[genes_to_use]
  EntrezID <- select(clariomdhumantranscriptcluster.db, IDs_to_use, c("ENTREZID"))
  EntrezID <- EntrezID$ENTREZID
  list_of_selected[[i]] <- EntrezID
  names(list_of_selected)[i] <- names(tables_to_use)[i]
}

sapply(list_of_selected, length)
````

Mapping GO and KEGG annotations:
````{r}
mapped_to_GO <- mappedkeys(org.Hs.egGO)
mapped_to_2KEGG <- mappedkeys(org.Hs.egPATH)
mapped_genes <- union(mapped_to_GO, mapped_to_2KEGG)
````

Applying Reactome in order to extract the results.
````{r}
list_of_data <- list_of_selected[1:3]
comparisons <- names(list_of_data)
universe <- mapped_genes

for (i in 1:length(list_of_data)){
  genes_IN <- list_of_data[[i]]
  comparison <- comparisons[i]
  enrich.result <- enrichPathway(gene = genes_IN,
                                 pvalueCutoff = 0.05,
                                 pAdjustMethod = "BH",
                                 organism = "human",
                                 universe = universe)
  if (length(rownames(enrich.result@result))!=0){
    write.csv2(as.data.frame(enrich.result),
               file = paste0("./resultats/","ReactomePA.results", comparison,".csv"),
               row.names = F)
    print(barplot(enrich.result, showCategory = 15, font.size = 4,
                  title = paste0(comparison)))
  }
}
````


